\large Neural and Symbolic Optimization of Cosmological Particle-Mesh Simulation \vspace{0.3em}

### 2023-09-07 docs/papers/sto/so.tex
FIXUP updating paper draft

In cosmological $N$-body simulations, the dynamics of the particles is governed
by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
In Fourier space, the gravitational force field $- \vnabla \varphi$ reads
%
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
%
We introduce force sharpening through
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$ and $g$ are NNs, each approximating a nonlinear function of the
wavevector components and all possible relevant features $\vvartheta$ summarized
in \autoref{tab:feat} and more details discussed in \autoref{sec:features}.
Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
broken by the PM solver to the O\textsubscript{h} discrete symmetry with
48 elements involving permutations and reversals of the 3 principal axes
of the simulation box.
The PM evaluation of each force component can further break the symmetry
along that axis, leaving the remaining 2 permutationally symmetric.
Therefore, we implement $f$ and $g$ to be even functions of the
wavevector components, and $g$ to be permutation equivariant in all 3
components, so that $f g$ is the most general modification to the force
kernel.
An even and real kernel in configuration space is also even in Fourier
space.
We take the absolute value for the evenness, and sort the 3 wavevector
compontents to achieve the permutation equivariance.
Being embedded in the differentiable simulation, the NNs can be trained using
gradient-based optimization methods as in usual deep learning tasks.
The parameters of the neural networks, which enter the force evaluation in
every time integration step, can be treated just the same as other cosmological
parameters.
The objective gradients are accumulated, based on which the parameters are then
optimized.
For both $f$ and $g$ neural networks, we adopt multilayer perceptron (MLP, i.e.
fully connected) structures.

@@

\subsection{Symbolic Expressions}

We use the random forest regressor for feature selection using impurity-based
importance.
\YZ{Maybe we can add another column in \autoref{tab:feat} for feature
importance?}
We use the \texttt{scikit-learn} \citep{scikit-learn} code for this.
% more details on the hyperparams in random forest

@@

We combine Lagrangian and Eulerian fields to form the total loss:
%
\begin{equation}
\cJ = \cL_\vs + [\cL_\vv] + \cL_\delta.
\end{equation}
%

@@

We have verified IC generation on GPUs is deterministic though different
from those on CPUs, so we generate ICs for both training data and
training on GPUs.

@@

@@

\YZ{Do we give this dataset a name and release it? This could be useful for
future training projects.}

@@

@@

We generate a sample of (input, output) data with the trained neural network to
perform feature selection and SR.

@@

\vspace{1em}
\textit{\large Code Availability:}
\pmwd\ \citep{Li2022b} is open-source on GitHub
\href{https://github.com/eelregit/pmwd}{\faGithub}, including the source
files and scripts of this paper
\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/sto}{\faFile}.

@@

>>>>

In cosmological $N$-body simulations, the dynamics of the particles is governed
by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
In Fourier space, the gravitational force field $- \vnabla \varphi$ reads
%
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
%
We introduce force sharpening through
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$ and $g$ are NNs, each approximating a nonlinear function of the
wavevector components and all possible relevant features $\vvartheta$ summarized
in \autoref{tab:feat} with more details discussed in \autoref{sec:features}.
Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
broken by the PM solver to the O\textsubscript{h} discrete symmetry with
48 elements involving permutations and reversals of the 3 principal axes
of the simulation box.
The PM evaluation of each force component can further break the symmetry
along that axis, leaving the remaining 2 permutationally symmetric.
Therefore, we implement $f$ and $g$ to be even functions of the
wavevector components, and $g$ to be permutation equivariant in all 3
components, so that $f g$ is the most general modification to the force
kernel.
An even and real kernel in configuration space is also even in Fourier
space.
We take the absolute value for the evenness, and sort the 3 wavevector
compontents to achieve the permutation equivariance.
Being embedded in the differentiable simulation, the NNs can be trained using
gradient-based optimization methods as in usual deep learning tasks.
The parameters of the neural networks, which enter the force evaluation in
every time integration step, can be treated just the same as other cosmological
parameters.
The objective gradients are accumulated, based on which the parameters are then
optimized.
For both $f$ and $g$ neural networks, we adopt multilayer perceptron (MLP, i.e.
fully connected) structures.

@@

\subsection{Feature Importance}
We use the random forest regressor for feature selection using impurity-based
importance.
\YZ{Maybe we can add another column in \autoref{tab:feat} for feature
importance?}

\subsection{Symbolic Expressions}
We use the \texttt{scikit-learn} \citep{scikit-learn} code for this.
% more details on the hyperparams in random forest

@@

We combine Lagrangian and Eulerian fields to form the total loss for a single
snapshot:
%
\begin{equation}
\cJ = \cL_\vs + \cL_\delta.
\end{equation}
%

@@

@@

Small numerical differences in initial conditions (ICs) could become more
significant with the $N$-body evolution.
We have verified that the IC generation on GPUs is deterministic though
different from those on CPUs, therefore we generate ICs for both the training
data and the model on GPUs.

@@

@@

\vspace{1em}
\subsection{Feature Importance Analysis}

We generate a sample of (input, output) data with the trained neural network to
perform feature importance analysis.

@@

@@

\vspace{1em}
\textit{\large Data Availability:}
\YZ{Do we give \GADGET\ Sobol 512 dataset a name and make it public? This could
be useful for future training projects.}

\vspace{1em}
\textit{\large Code Availability:}
\pmwd\ \citep{Li2022b} is open-source on GitHub
\href{https://github.com/eelregit/pmwd}{\faGithub}, including the source files
and scripts of this paper
\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/sto}{\faFile}.
\GADGET\ \citep{GADGET-4}, which can be obtained through a public git-repository
\href{http://gitlab.mpcdf.mpg.de/vrs/gadget4}{\faGitlab}, hosted by a gitlab
server at the Max-Planck Computing and Data Facility (MPCDF).

@@

\appendix
% supplementary information of Nat. Mach. Intell.
